---
title: 'Quantitative evidence synthesis for biodiversity'
author: "Shane Blowes"
date: "Shanghai 2024"
output: 
  html_document:
    code_folding: hide
bibliography: references2.bib  
---
  
The use of quantitative evidence synthesis has accelerated exponentially since
approximately the mid-1990s [@gurevitch2018meta].  

![](/Users/sb25gaqy/Dropbox/4teaching/shanghai-biodiversity-synthesis-course-2024/Day 4 August 31 How to ask questions and develop hypotheses (for biodiversity synthesis)/meta-analysis and biodiversity change synthesis.png)   


This broad trend for quantitative evidence synthesis, and meta-analysis in particular, 
is also true for ecologists and conservation biologists [@anderson2021trends].

![](/Users/sb25gaqy/Dropbox/4teaching/shanghai-biodiversity-synthesis-course-2024/Day 4 August 31 How to ask questions and develop hypotheses (for biodiversity synthesis)/evidence-synthesis-models-and-methods.png)   

For biodiversity synthesis, two obstacles soon become apparent 
when we start diving into the literature search for a meta-analysis. 

First, the tyranny of the many indices will quickly become apparent: 
no matter which effect size you target
(e.g., log-ratio, Hedge's G, etc), the many different biodiversity measures (e.g.,
species richness, total abundance, Shannon diversity, Simpson diversity,
one of the Hill numbers, etc.) will limit 
the number of studies you can collate. Second, the two things that we emphasise
as being important for describing variation in biodiversity - scale and the 
multiple components of abundance, evenness and richness - will be inaccessible
for most meta-analyses. Multi-scale analyses are predominantly out of reach 
because most studies will have only quantified biodiversity at a single scale. 
Similarly, multi-component analyses will be constrained by the fact that most studies 
will have used one or two metrics, and the scope for coherently combining these 
metrics across studies will be limited.

Even for a single metric such as species richness, a within system meta-analysis
will risk comparing apples to oranges because different studies will likely have
used different sample grains to sample the target assemblage. Let's do a simple
simulation to examine the implication of variation in grain size for our ability 
to generalise an effect size.

### How can scale, e.g., grain size, influence a meta-analysis? 

To examine how variation in grain size influences a meta-analysis, we will use
simulations. Let's start by simulating an experiment that removes half of the 
species in a community. This is similar to the experiment we did with the candy,
but now we will have the treatment remove species instead of individuals. There
are a few common effect sizes used in meta-analysis, here we'll use the log-ratio
[@hedges1999meta].

```{r, warning=FALSE, class.source = 'fold.show', message=FALSE}
library(tidyverse)
library(mobsim)
library(mobr)
library(cowplot)

# set a RNG seed 
set.seed(42)

# Set all communities to have 5000 individuals, 
# a lognormal SAD, and individuals randomly distributed in space
Jpool <- 2000

# We want to treatment to half the number of species
Spool_control <- 200
# treatment removes 50% of the species
Spool_treatment <- 0.5 * Spool_control

# simulate a meta-analysis with twenty studies
n_study <- 20

meta_sim <- tibble(
  Jpool = Jpool,
  S_control = Spool_control,
  S_treatment = Spool_treatment) %>% 
  # create the n_studies
  uncount(n_study, .remove = FALSE) %>% 
  # here, we are interested in examining variation in sample grain.
  # Draw some random quadrat sizes from a uniform distribution
  mutate(sample_grain = runif(n_study,
                              min = 0.01,
                              max = 0.1)) %>% 
  # create index to identify each study
  mutate(study = as.character(1:n_study)) %>% 
  # prepare to generate control and treatment communities for each study
  group_by(study) %>% 
  nest(data = c(Jpool, S_control, S_treatment,
                sample_grain)) %>% 
  # simulate poisson distribution of individuals for the controls and 
  # treatments within each study
  mutate(control_comm = map(data, 
                            ~sim_poisson_community(
                              s_pool = .x$S_control,
                              n_sim = .x$Jpool,
                              sad_type = 'lnorm',
                              sad_coef = list('meanlog' = log(.x$S_control/.x$Jpool),
                                              'sdlog' = 1))),
         treatment_comm = map(data, 
                              ~sim_poisson_community(
                                s_pool = .x$S_treatment,
                                n_sim = .x$Jpool,
                                sad_type = 'lnorm',
                                sad_coef = list('meanlog' = log(.x$S_treatment/.x$Jpool),
                                                'sdlog' = 1)
                                )
                              )
         ) %>% 
  # to check our treatment worked calculate abundance and richness for the 
  # whole community (which is the scale at which we applied the treatment)
  mutate(control_comm_J = map(control_comm, ~nrow(.x$census)),
         control_comm_S = map(control_comm, ~n_distinct(.x$census$species)),
         treatment_comm_J = map(treatment_comm, ~nrow(.x$census)),
         treatment_comm_S = map(treatment_comm, ~n_distinct(.x$census$species))) %>% 
  # now, we want to get some samples from the controls and treatments
  # we'll just keep the site x species matrix
  mutate(control_samps = map2(control_comm, data, 
                             ~ sample_quadrats(comm = .x,
                                               # we'll take 5 samples
                                               n_quadrats = 5,
                                               # with the sample_grain for this study
                                               quadrat_area = .y$sample_grain,
                                               method = 'grid',
                                               plot = FALSE)$spec_dat),
         treatment_samps = map2(treatment_comm, data, 
                             ~ sample_quadrats(comm = .x,
                                               # we'll take 5 samples
                                               n_quadrats = 5,
                                               # with the sample_grain for this study
                                               quadrat_area = .y$sample_grain,
                                               method = 'grid',
                                               plot = FALSE)$spec_dat)) %>% 
  # let's calculate metrics for abundance, richness, and the ENS conversion
  # of PIE (for inferences about evenness)
  # note, because we're simulating what we'd typically find when doing a meta-analysis,
  # we're only going to do the calculations at a single scale
  mutate(control_J = map(control_samps, ~ rowSums(.x)),
         control_S = map(control_samps, ~ vegan::specnumber(.x, MARGIN = 1)),
         control_SPIE = map(control_samps, ~ mobr::calc_comm_div(.x, index = 'S_PIE', 
                                                                extrapolate = FALSE,
                                                                scales = 'alpha')$value),
         treatment_J = map(treatment_samps, ~ rowSums(.x)),
         treatment_S = map(treatment_samps, ~ vegan::specnumber(.x, MARGIN = 1)),
         treatment_SPIE = map(treatment_samps, ~ mobr::calc_comm_div(.x, index = 'S_PIE', 
                                                                extrapolate = FALSE,
                                                                scales = 'alpha')$value)) %>% 
  ungroup()
  
# calculate log-ratio effect sizes
effect_sizes <- meta_sim %>% 
  unnest(c(data, control_J, control_S, control_SPIE,
           treatment_J, treatment_S, treatment_SPIE)) %>% 
  mutate(J_LRR = log(treatment_J/control_J),
         S_LRR = log(treatment_S/control_S),
         SPIE_LRR = log(treatment_SPIE/control_SPIE))

```

We're not focused on the details of meta-analysis today, so we won't fit any
statistical models. Instead, we'll look at the data and calculate 
some averages. 

```{r, warning=FALSE, class.source = 'fold.show', message=FALSE, fig.dim = c(6,4)}

# plot come effect sizes, let's start with total numbers of individuals
effect_sizes %>% 
  # calculate the mean and sd
  mutate(J_LRR_mean = mean(J_LRR),
         J_LRR_sd = sd(J_LRR)) %>% 
  ggplot() +
  geom_rect(aes(xmin = -Inf, xmax = Inf,
                  ymin = J_LRR_mean - J_LRR_sd,
                  ymax = J_LRR_mean + J_LRR_sd),
                fill = 'light grey') +
  geom_hline(aes(yintercept = J_LRR_mean)) +
  # and plot the known effect size on total abundance
  geom_hline(yintercept = 0, lty = 2) +
  geom_point(aes(x = study, y = J_LRR)) +
  labs(x = 'Study',
       y = 'Effect size (LRR)',
       subtitle = 'Treatment effect on total abundance') +
  theme_bw()
```

Nice. We have got the no effect on total abundance approximately 
right.

What about species richness?

```{r, warning=FALSE, class.source = 'fold.show', message=FALSE, fig.dim = c(6,4)}
# now species richness, note our known effect size
S_LRR_known <- log(0.5)

effect_sizes %>% 
  # calculate the mean and sd
  mutate(S_LRR_mean = mean(S_LRR),
         S_LRR_sd = sd(S_LRR)) %>% 
  ggplot() +
  geom_rect(aes(xmin = -Inf, xmax = Inf,
                  ymin = S_LRR_mean - S_LRR_sd,
                  ymax = S_LRR_mean + S_LRR_sd),
                fill = 'light grey') +
  geom_hline(aes(yintercept = S_LRR_mean)) +
  # let's plot the known effect size as a dashed line
  geom_hline(yintercept = S_LRR_known, linetype = 2) + 
  geom_point(aes(x = study, y = S_LRR)) +
  labs(x = 'Study',
       y = 'Effect size (LRR)',
       subtitle = 'Treatment effect on species richness') +
  theme_bw()

```

Here, it looks like we got the sign (direction) of the effect right. But we have 
underestimated the effect (sometimes called a magnitude error). The magnitude, i.e.,
how much we under or overestimate the effect, would depend on the distribution 
of grain sizes, and the scale at which the
effect impacted biodiversity, both of which would likely vary among studies. 
Here we know that the effect was applied at a grain size 
of one, while our grain sizes between 0.01 and 0.1. In an empirical synthesis of
observational data, we would not know the true effect (nor the scale at which
it would impact the community).

An important recommendation for all biodiversity syntheses, and in particular, 
for meta-analyses with a diversity measure as the response variable is to plot
effect sizes as a function of scale (e.g., sample grain and extent) 
[@spake_implications_2021].

```{r, warning=FALSE, class.source = 'fold.show', message=FALSE, fig.dim = c(6,4)}
effect_sizes %>% 
  # calculate the mean and sd of the sample ESs
  mutate(S_LRR_mean = mean(S_LRR),
         S_LRR_sd = sd(S_LRR)) %>% 
  ggplot() +
  geom_hline(aes(yintercept = S_LRR_mean)) +
  # plot the known effect size as a dashed line
  geom_hline(yintercept = S_LRR_known, linetype = 2) + 
  # plot the sample effect sizes
  geom_point(aes(x = sample_grain, y = S_LRR)) +
  # to see if we've got any evidence for scale-dependence
  # we'll visualise a linear model with the effect sizes as a function of
  # grain size
  stat_smooth(aes(x = sample_grain, y = S_LRR),
              method = 'lm') +
  labs(x = 'Grain size',
       y = 'Effect size (LRR)',
       subtitle = 'Estimated treatment effect on species richness') +
  theme_bw()

```

Some grain-size dependence in the average effect size for species richness. And
larger grains are closer to the known ES.

Let's repeat the exercise, but this time we'll have the treatment effect the
numbers of individuals. This is exactly what we tried to do with the candy 
experiment, but now it will be much easier to randomly remove individuals.
Before we start, what is your expectation? Do you think
we'll be able to accurately recover the known effect size for total numbers of 
individuals independently of the grain size variation? What about species richness?

```{r, warning=FALSE, class.source = 'fold.show', message=FALSE}
# this time both control and treatment have the same number of species
Spool <- 200

# Simulate different communities sizes for the control and treatment
Jpool_control <- 5000
# treatment removes 60% of the individuals
Jpool_treatment <- 2000


meta_sim2 <- tibble(
  Spool = Spool,
  J_control = Jpool_control,
  J_treatment = Jpool_treatment,
) %>% 
  # create the n_studies
  uncount(n_study, .remove = FALSE) %>% 
  # here, we are interested in examining how variation in sample grain.
  # So, we'll draw some random quadrat sizes from a uniform distribution
  mutate(sample_grain = runif(n_study,
                              min = 0.01,
                              max = 0.1)) %>% 
  # create index to identify each study
  mutate(study = 1:n_study) %>% 
  # prepare to generate control and treatment communities for each study
  group_by(study) %>% 
  nest(data = c(Spool, J_control, J_treatment,
                sample_grain)) %>% 
  # simulate poisson distribution of individuals for the controls and 
  # treatments within each study
  mutate(control_comm = map(data, ~ sim_poisson_community(s_pool = .x$Spool,
                                          n_sim = .x$J_control,
                                          sad_type = 'lnorm',
                                          sad_coef = list('meanlog' = log(.x$Spool/.x$J_control),
                                                          'sdlog' = 1))),
         treatment_comm = map(data, ~ sim_poisson_community(s_pool = .x$Spool,
                                          n_sim = .x$J_treatment,
                                          sad_type = 'lnorm',
                                          sad_coef = list('meanlog' = log(.x$Spool/.x$J_treatment),
                                                          'sdlog' = 1)))) %>% 
  # now, we want to get some samples from the controls and treatments
  # we'll just keep the site x species matrix
  mutate(control_samps = map2(control_comm, data, 
                             ~ sample_quadrats(comm = .x,
                                               # we'll take 5 samples
                                               n_quadrats = 5,
                                               # with the sample_grain for this study
                                               quadrat_area = .y$sample_grain,
                                               method = 'grid',
                                               plot = FALSE)$spec_dat),
         treatment_samps = map2(treatment_comm, data, 
                             ~ sample_quadrats(comm = .x,
                                               # we'll take 5 samples
                                               n_quadrats = 5,
                                               # with the sample_grain for this study
                                               quadrat_area = .y$sample_grain,
                                               method = 'grid',
                                               plot = FALSE)$spec_dat)) %>% 
  # let's calculate metrics for abundance, richness, and the ENS conversion
  # of PIE (for inferences about evenness)
  # note, because we're simulating what we'd typically find when doing a meta-analysis,
  # we're only going to do the calculations at a single scale
  mutate(control_J = map(control_samps, ~ rowSums(.x)),
         control_S = map(control_samps, ~ vegan::specnumber(.x, MARGIN = 1)),
         control_SPIE = map(control_samps, ~ mobr::calc_comm_div(.x, index = 'S_PIE', 
                                                                extrapolate = FALSE,
                                                                scales = 'alpha')$value),
         treatment_J = map(treatment_samps, ~ rowSums(.x)),
         treatment_S = map(treatment_samps, ~ vegan::specnumber(.x, MARGIN = 1)),
         treatment_SPIE = map(treatment_samps, ~ mobr::calc_comm_div(.x, index = 'S_PIE', 
                                                                extrapolate = FALSE,
                                                                scales = 'alpha')$value)) %>% 
  ungroup()
  
# calculate log-ratio effect sizes
effect_sizes2 <- meta_sim2 %>% 
  unnest(c(data, control_J, control_S, control_SPIE,
           treatment_J, treatment_S, treatment_SPIE)) %>% 
  mutate(J_LRR = log(treatment_J/control_J),
         S_LRR = log(treatment_S/control_S),
         SPIE_LRR = log(treatment_SPIE/control_SPIE))
```

Again, we'll focus on visual inspection and the averages instead of fitting
statistical models and a more formal analysis.

```{r, warning=FALSE, class.source = 'fold.show', message=FALSE, fig.dim = c(6,4)}

# plot come effect sizes, let's start with total numbers of individuals
J_LRR_known = log(2/5)
  
effect_sizes2 %>% 
  # calculate the mean and sd
  mutate(J_LRR_mean = mean(J_LRR),
         J_LRR_sd = sd(J_LRR)) %>% 
  ggplot() +
  geom_rect(aes(xmin = -Inf, xmax = Inf,
                  ymin = J_LRR_mean - J_LRR_sd,
                  ymax = J_LRR_mean + J_LRR_sd),
                fill = 'light grey') +
  geom_hline(aes(yintercept = J_LRR_mean)) +
  # and plot the known effect size on total abundance
  geom_hline(yintercept = J_LRR_known, lty = 2) +
  geom_point(aes(x = study, y = J_LRR)) +
  labs(x = 'Study',
       y = 'Effect size (LRR)',
       subtitle = 'Estimated treatment effect on total abundance') +
  theme_bw()
```

We still estimate the known effect on total numbers of individuals well. Why?

What about species richness?

```{r, warning=FALSE, class.source = 'fold.show', message=FALSE, fig.dim = c(6,4)}
# now species richness, note our known effect size
S_LRR_known <- 0

effect_sizes2 %>% 
  # calculate the mean and sd
  mutate(S_LRR_mean = mean(S_LRR),
         S_LRR_sd = sd(S_LRR)) %>% 
  ggplot() +
  geom_rect(aes(xmin = -Inf, xmax = Inf,
                  ymin = S_LRR_mean - S_LRR_sd,
                  ymax = S_LRR_mean + S_LRR_sd),
                fill = 'light grey') +
  geom_hline(aes(yintercept = S_LRR_mean)) +
  # let's plot the known effect size as a dashed line
  geom_hline(yintercept = S_LRR_known, linetype = 2) + 
  geom_point(aes(x = study, y = S_LRR)) +
  labs(x = 'Study',
       y = 'Effect size (LRR)',
       subtitle = 'Estimated treatment effect on species richness') +
  theme_bw()

```

Now we've overestimated the effect on species richness! Again, the magnitude of our 
error would depend on the distribution of grain sizes, and the scale at which the
effect impacted biodiversity (both of which would be unknown in an empirical 
meta-analysis, though here we know that there was no effect on community richness). 

Let's look at whether we have strong scale dependence in the effect sizes
[@spake_implications_2021].

```{r, warning=FALSE, class.source = 'fold.show', message=FALSE, fig.dim = c(6,4)}
effect_sizes2 %>% 
  # calculate the mean and sd
  mutate(S_LRR_mean = mean(S_LRR),
         S_LRR_sd = sd(S_LRR)) %>% 
  ggplot() +
  geom_hline(aes(yintercept = S_LRR_mean)) +
  # let's plot the known effect size as a dashed line
  geom_hline(yintercept = S_LRR_known, linetype = 2) + 
  geom_point(aes(x = sample_grain, y = S_LRR)) +
  # let's see if we've got any evidence for scale-dependence
  # by fitting a linear model to the effect size as a function of
  # grain size
  stat_smooth(aes(x = sample_grain, y = S_LRR),
              method = 'lm') +
  labs(x = 'Grain size',
       y = 'Effect size (LRR)',
       subtitle = 'Estimated treatment effect on species richness') +
  theme_bw()

```

Our effect size estimates are larger at smaller scales, and decrease with
increasing grain size. Again, we get 
closer to the known ES with larger grains. Why?

What happened to evenness? Is our estimated ES grain-size dependent?

```{r, warning=FALSE, class.source = 'fold.show', message=FALSE, fig.dim = c(6,4)}
effect_sizes2 %>% 
  # calculate the mean and sd
  mutate(SPIE_LRR_mean = mean(SPIE_LRR),
         SPIE_LRR_sd = sd(SPIE_LRR)) %>% 
  ggplot() +
  geom_hline(aes(yintercept = SPIE_LRR_mean)) +
  # let's plot the known effect size as a dashed line
  geom_hline(yintercept = 0, linetype = 2) + 
  geom_point(aes(x = sample_grain, y = SPIE_LRR)) +
  # let's see if we've got any evidence for scale-dependence
  # by fitting a linear model to the effect size as a function of
  # grain size
  stat_smooth(aes(x = sample_grain, y = SPIE_LRR),
              method = 'lm') +
  labs(x = 'Grain size',
       y = 'Effect size (LRR)',
       subtitle = 'Estimated treatment effect on evenness') +
  theme_bw()
```

### Exercise: What about within-species aggregation?

Our initial simulations had communities where individuals were distributed 
randomly in space. This helped simplify interpreting what we found, but is 
likely unrealistic for most communities (though it might be appropriate for
some experimental conditions). Repeat the simulations, but this time add some
within species aggregation.

# Can we solve any of these problems?

Variation in grain size can limit our ability to do biodiversity
synthesis. This is especially true when we attempt to synthesise information 
using meta-analysis. But what about if we can get the **raw** data? In this
context, raw data is the abundance of all 
the species in a sample, as opposed to a single number (i.e., a diversity measure
or index) that summarises the sample (like species richness or Shannon diversity).

Let's first have a look at a Species Abundance Distribution (SAD).

```{r, warning=FALSE, class.source = 'fold.show', message=FALSE}
sim_sad(s_pool = 10, n_sim = 200, sad_type = 'lnorm', 
        sad_coef = list('meanlog' = log(10/200),
                        'sdlog' = 1))
```

The key here is that we have the abundance of each species. Not a single number
summarising the diversity for the whole sample.
With this information - counts of individuals are ideal - we can use 
rarefaction to do a better comparison between samples with different grains. 
Specifically, we will standardise our samples to have the same number of individuals.
And then compare the expected diversity for a common (standardised) number of 
individuals. Importantly, this standardisation can only be applied directly at the
alpha-scale (where we are standardising the sampling grain). Additional 
work is required at the gamma-scale, where variation in extent can also confound
comparisons.

Again, we'll use some simple simulations to illustrate how standardisation by 
numbers of individuals can work. Using the data from our earlier simulated 
experiment that removed species, we will standardise the sample
effort before we calculate effect sizes.

```{r, warning=FALSE, class.source = 'fold.show', message=FALSE}

# For this exercise, we'll reuse the data we simulated earlier. But, now we want
# use the SAD, i.e., the raw data from the sample, to standardise sampling effort
# before we calculate the effect sizes. 

# we want to work with the control_samps and treatment_samps
effort_standardisation <-
meta_sim %>% 
  select(study, data, control_samps, treatment_samps) %>% 
  # need long data for rarefaction
  mutate(control_long = map(control_samps, 
                            ~rownames_to_column(., var = 'site') %>% 
                              pivot_longer(cols = !site,
                                           names_to = 'species',
                                           values_to = 'N')),
         treatment_long = map(treatment_samps, 
                            ~rownames_to_column(., var = 'site') %>% 
                              pivot_longer(cols = !site,
                                           names_to = 'species',
                                           values_to = 'N'))) 
  
# let's separate out the control and treatment samples
# calculate the IBR for each site
control_long <- effort_standardisation %>% 
  select(study, control_long) %>%
  unnest(c(control_long)) %>% 
  group_by(study, site) %>% 
  nest(site_data = c(species, N)) %>% 
  mutate(expected_richness = map(site_data, ~rarefaction(.x$N, method = 'IBR'))) %>% 
  unnest(c(expected_richness)) %>% 
  mutate(individuals = 1:n()) %>% 
  ungroup() %>% 
  # put the grain sizes back in
  left_join(meta_sim %>% 
              select(study, data) %>% 
              unnest(c(data))) %>% 
  mutate(treatment = 'control')

treatment_long <- effort_standardisation %>% 
  select(study, treatment_long) %>% 
  unnest(c(treatment_long)) %>% 
  group_by(study, site) %>% 
  nest(site_data = c(species, N)) %>% 
  mutate(expected_richness = map(site_data, ~rarefaction(.x$N, method = 'IBR'))) %>% 
  unnest(c(expected_richness)) %>% 
  mutate(individuals = 1:n()) %>% 
  ungroup() %>% 
  # put the grain sizes back in
  left_join(meta_sim %>% 
              select(study, data) %>% 
              unnest(c(data))) %>% 
  mutate(treatment = 'removal')
```

First, let's use rarefaction curves to visualise the samples. And we'll add a 
vertical line showing the minimum number of individuals across all the samples.

```{r, warning=FALSE, class.source = 'fold.show', message=FALSE, fig.dim(c(8,4))}
# join and plot
bind_rows(control_long, 
          treatment_long) %>% 
  ggplot() +
  facet_wrap(~treatment) +
  geom_line(aes(x = individuals, y = expected_richness,
                group = interaction(study, site), colour = sample_grain)) +
  # add a vertical line with the minimum number of individuals across all of the
  # samples
  geom_vline(data = . %>% 
               # first find the number of individuals in each sample
               group_by(study, site, treatment) %>% 
               filter(individuals == max(individuals)) %>% 
               ungroup() %>% 
               summarise(minJ = min(individuals)),
             aes(xintercept = minJ), lty = 2)

```

Here, we are interested in rarefying to the number of individuals associated 
with our smallest grain size. We can use this because total abundance scales
approximately linearly with grain size.

```{r, warning=FALSE, class.source = 'fold.show', message=FALSE}
# plot mean J (of our samples) as a function of grain
# total abundance scales approximately linearly
bind_rows(control_long,
          treatment_long) %>% 
  group_by(site, sample_grain) %>% 
  summarise(J = n()) %>% 
  group_by(sample_grain) %>% 
  summarise(Jbar = mean(J)) %>% 
  ungroup() %>% 
  ggplot() + 
  geom_point(aes(x = sample_grain, y = Jbar)) +
  stat_smooth(aes(x = sample_grain, y = Jbar),
              method = 'gam') +
  labs(x = 'Sample grain',
       y = 'Average number of individuals')

# want to get average abundance in the samples with the smallest grain.
# this will become our target number of individuals that we will interpolate
# the other samples to
target_J = bind_rows(control_long,
          treatment_long) %>% 
  filter(sample_grain == min(sample_grain)) %>% 
  group_by(study, site, treatment, sample_grain) %>% 
  summarise(J = n()) %>% 
  group_by(sample_grain) %>% 
  summarise(Jbar = round(mean(J))) 

# now calculate expected richness for target_J
effort_standardisation <- effort_standardisation %>% 
  mutate(control_expectedS = map(control_samps, ~vegan::rarefy(., sample = target_J)[,2]),
         treatment_expectedS = map(treatment_samps, ~vegan::rarefy(., sample = target_J)[,2]))

standardised_effect_sizes <- effort_standardisation %>% 
  unnest(c(data, control_expectedS, treatment_expectedS)) %>% 
  mutate(expected_S_LRR = log(treatment_expectedS/control_expectedS))

standardised_effect_sizes %>% 
  # calculate the mean and sd of the effort-standardised effect sizes
  mutate(LRR_mean = mean(expected_S_LRR),
         LRR_sd = sd(expected_S_LRR)) %>% 
  ggplot() +
  geom_hline(aes(yintercept = LRR_mean)) +
  # plot the effort-standardised effect sizes
  geom_point(aes(x = sample_grain, y = expected_S_LRR)) +
  # to see if we've got any evidence for scale-dependence
  # we'll visualise a linear model with the effect sizes as a function of
  # grain size
  stat_smooth(aes(x = sample_grain, y = expected_S_LRR),
              method = 'lm') +
  labs(x = 'Grain size',
       y = 'Effect size (LRR)',
       subtitle = 'Estimated treatment effect on effort-standardised species richness') +
  theme_bw()

```

We still underestimate the known effect of the treatment on species richness. 
This is because the grain of our samples is small compared to the scale at 
which the effect modified species richness.

However, we have at largely removed the grain size dependence of the effect size.
And the variation in the effect sizes that remains, can more reasonably be compared.
i.e., some studies had a smaller effect than average, others had a larger effect
than the average.
effect sizes by standardising sample effort 
### References
